Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-06 source/paper.tex

\JScomment{This paragraph should refer to the schematic and explain
  all aspects} The Gusto model comprises a domain and a prognostic
equation set. The domain holds the mesh and the finite element
function spaces along with any other information about the space-time
domain such as the coordinates and the timestep. The prognostic
equation sets are each defined in a python class, making use of the
common finite element forms for the transport and difffusion terms to
set up the residual (defined in section \ref{}) in weak form. At this
point, no particular discretisation of the transport or diffusion
forms is specified. This happens in the \texttt{SpatialMethod} class
which can be used to apply the methods described in section \ref{} to
the relevant terms. Any physics source terms, such as **, are added by
calling the \texttt{Physics} class that defines them. The
\texttt{TimeStepper} class contains the definition of the timestepping
procedure and the \texttt{run} method that loops over the
timesteps. It also sets up diagnostic fields and calls the IO as
required. The simplest timeloop class will just apply one time
discretisation method to the equation set but it is common in weather
and climate prediction models to split the different types of terms
and apply a time discretisation separately to each type of term. For
example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
involved in physics parameterisations are evaluated in the appropriate
part of the iteration loop, with physics processes that occur on
`fast' timescales evaluated more frequently and `slower' physics
processes evaluated just once at the end of the timestep. The choice
of where to evaluate the physics processes in the iteration is also
based on the computational cost and the best way to do this is still
an open research question (**ref).

\begin{itemize}
\item what about initialisation and diagnostics?
\end{itemize}

\begin{algorithm}
  \caption{Pseudocode for SIQN timestep loop}\label{alg:SIQN}
  \begin{algorithmic}
    \STATE
  \end{algorithmic}
\end{algorithm}

\begin{itemize}
\item generic example(s) - no FML, just pseudocode for basic
  timestepper, SIQN and split physics
\item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
\end{itemize}

@@

\subsection{Compressible Euler equations}
We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern weather and climate models.
In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
 \\
\\
The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
Figure \ref{fig:straka} shows the final potential temperature field for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.8\textwidth]{figures/straka.png}
\caption{The potential temperature field after 900 s from the falling density current test of \citet{straka1993numerical}. (Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
}
\label{fig:straka}
\end{figure}
\\
To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
The initial conditions are those of \citet{bendall2020compatible}, which is an extension of the dry case of \citet{kelly2012continuous} and \citet{melvin2019mixed}.
A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
As in the two-dimensional test of \citet{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
A simple saturation adjustment parametrisation is used, with the details described in \citet{bendall2020compatible}.
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.48\textwidth]{figures/moist_3d_bubble.png}
\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
}
\label{fig:3d_bubble}
\end{figure}
\\
Our final test case is the deep baroclinic wave of \citet{ullrich2014proposed} on the sphere.
An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in numerical weather prediction.
The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citet{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.

>>>>

\JScomment{This paragraph should refer to the schematic and explain
  all aspects} The Gusto model comprises a domain and a prognostic
equation set. The domain holds the mesh and the finite element
function spaces along with any other information about the space-time
domain such as the coordinates and the timestep. The prognostic
equation sets are each defined in a python class, making use of the
common finite element forms for the transport and diffusion terms to
set up the residual (i.e. the left hand side of the equations defined
in section \ref{sec: equations}) in weak form. At this point, no
particular discretisation of the transport or diffusion forms is
specified. This happens in the \texttt{SpatialMethod} class which can
be used to apply the methods described in sections \ref{sec:
  transport} and \ref{sec: diffusion} to the relevant terms. Any
physics source terms, such as those required for moist processes or
tubulence parameterisation, are added by calling the \texttt{Physics}
class that defines them. The \texttt{TimeStepper} class contains the
definition of the timestepping procedure and the \texttt{run} method
that loops over the timesteps. It also sets up diagnostic fields and
calls the IO as required. The simplest timeloop class will just apply
one time discretisation method to the equation set but it is common in
weather and climate prediction models to split the different types of
terms and apply a time discretisation separately to each type of
term. For example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
involved in physics parameterisations are evaluated in the appropriate
part of the iteration loop, with physics processes that occur on
`fast' timescales evaluated more frequently and `slower' physics
processes evaluated just once at the end of the timestep. Since the
test cases presented later only use simple moist physics
parameterisations, we just include the option to solve for these at
the end of the timestep.

\begin{algorithm} 
  \caption{Basic timestepping scheme}
  \label{alg: basic}
  \begin{algorithmic}[1]
    \STATE Replace transport and diffusion terms with specified spatial discretisation scheme.
    \STATE Setup IO and diagnostics and write out initial conditions.
    \WHILE{$t < t_{max}$}
      \STATE Apply timestepping scheme to equation set: $\chi^{n+1} = \mathcal{T}(\chi^n, \chi^{n+1})$
      \STATE Advance timestep: $\chi^n = \chi^{n+1}$
      \IF{ndt \% nts == 0}
        \STATE Update timeseries diagnostics file.
      \ENDIF
      \IF{ndt \% ndiag == 0}
        \STATE Compute and write out diagnostic fields.
      \ENDIF
      \IF{ndt \% nchk == 0}
        \STATE Write out checkpoint file.
      \ENDIF
    \ENDWHILE
  \end{algorithmic} 
\end{algorithm}

\begin{algorithm} 
  \caption{Split physics timestepping scheme}
  \label{alg: split physics}
  \begin{algorithmic}[1]
    \STATE Dynamics step: $\chi^{n+1} = \mathcal{D}(\chi^n)$
    \FOR {$\mathcal{P}$ in physics schemes}
      \STATE Apply physics scheme: $\chi^{n+1} = \chi^{n+1} + \mathcal{P}(\chi^{n+1})$
    \ENDFOR
    \STATE Advance timestep: $\chi^n = \chi^{n+1}$
  \end{algorithmic} 
\end{algorithm}

\begin{algorithm} 
  \caption{Semi-implicit quasi-Newton timestepping scheme}
  \label{alg: SIQN}
  \begin{algorithmic}[1]
    \STATE Set: $\chi^{n+1} = \chi^{n}$
    \STATE Explicit forcing: $\chi_{f, e} = \chi^n + \frac{1}{2}\Delta t \mathcal{F}(\chi^n)$
    \FOR {$OUTER$}
      \STATE Update advecting velocity: $\bar{\MM{u}} = \frac{1}{2} (\MM{u}^{n+1} + \MM{u}^n)$
      \STATE Explicit transport: $\chi_T = \mathcal{A}_{\bar{\MM{u}}}(\chi_{f,e})$
      \FOR {$INNER$}
        \STATE Implicit forcing: $\chi_{f,i} = \chi_T + \frac{1}{2}\Delta t \mathcal{F}(\chi^{n+1})$
        \STATE Write residual: $\chi_{r} = \chi_{f,i} - \chi^{n+1}$
        \STATE Solve: $\mathcal{S}(\chi') = \chi_{r}$ for $\chi'$
        \STATE Increment: $\chi^{n+1} = \chi^{n+1} + \chi'$
      \ENDFOR
    \ENDFOR
    \STATE Explicit physics: $\chi^{n+1} = \mathcal{P}(\chi^{n+1})$
    \STATE Advance timestep: $\chi^n = \chi^{n+1}$
  \end{algorithmic} 
\end{algorithm}

@@

\subsection{Compressible Euler equations}
We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern weather and climate models.
In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
 \\
\\
The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
Figure \ref{fig:straka} shows the final potential temperature field for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.6\textwidth]{figures/straka.png}
\caption{The potential temperature field after 900 s from the falling density current test of \citet{straka1993numerical}. (Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
}
\label{fig:straka}
\end{figure}
\\
To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
The initial conditions are those of \citet{bendall2020compatible}, which is an extension of the dry case of \citet{kelly2012continuous} and \citet{melvin2019mixed}.
A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
As in the two-dimensional test of \citet{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
A simple saturation adjustment parametrisation is used, with the details described in \citet{bendall2020compatible}.
\\
\begin{figure}[htp!]
\centering
\includegraphics[width=0.48\textwidth]{figures/moist_3d_bubble.png}
\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
}
\label{fig:3d_bubble}
\end{figure}
\\
Our final test case is the deep baroclinic wave of \citet{ullrich2014proposed} on the sphere.
An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in numerical weather prediction.
The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citet{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
